---
title: 'Lecture Notes on the Politics of Information, Communication, and Technology'
author: 'Justin Murphy'
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: yes
    theme: readable
    keep_md: yes
  pdf_document:
    fig_caption: yes
    includes:
      in_header: mystyles.sty
    keep_tex: yes
    latex_engine: xelatex
    toc: yes
citecolor: red
fontsize: 12pt
geometry: margin=1.5in
linkcolor: red
mainfont: PT Serif
classoption: book
bibliography: mybib.bib
urlcolor: red
---

Outline

\pagebreak


```{r, echo=FALSE}
setwd("~/Dropbox/gh_projects/pacification_of_the_world")
```

# Part 1: General Theoretical Reflections
```{r general-theory, child="~/Dropbox/gh_projects/pacification_of_the_world/content/general_theory.md"}
```

## The conservatism of everyday speech
```{r isntrumentalism-of-speech, child="~/Dropbox/gh_projects/pacification_of_the_world/content/conservatism_of_everyday_speech.md"}
```

## Democracy as propaganda, 1914-1919
```{r propaganda_and_democracy, child="~/Dropbox/gh_projects/pacification_of_the_world/content/propaganda_and_democracy/propaganda_and_democracy.Rmd"}
```

# The emergence of public opinion
```{r creation_of_public_opinion, child="~/Dropbox/gh_projects/pacification_of_the_world/content/creation_of_public_opinion.md"}
```

# On Walter Lippmann
```{r on_lippmann, child="~/Dropbox/gh_projects/pacification_of_the_world/content/on_lippmann/on_lippmann.Rmd"}
```

# On Bell Labs
```{r bell_labs_information_theory, child="~/Dropbox/gh_projects/pacification_of_the_world/content/bell_labs_information_theory.md"}
```

# A distributive theory of information processing
```{r model_digital_political, child="~/Dropbox/gh_projects/pacification_of_the_world/content/model_digital_political.md"}
```

# On the early social science of the media
```{r early_science_of_media, child="~/Dropbox/gh_projects/pacification_of_the_world/content/early_science_of_media.md"}
```

# Social science is endogenous to the media
```{r social_science_is_endogenous, child="~/Dropbox/gh_projects/pacification_of_the_world/content/social_science_is_endogenous/social_science_is_endogenous.Rmd"}
```

# The ideology of "limited effects" in the institutionalized social sciences
```{r ideology_of_limited_effects, child="~/Dropbox/gh_projects/pacification_of_the_world/content/ideology_of_limited_effects.md"}
```

# Old but unresolved problems with public opinion
```{r problems_with_the_opinion_policy_link, child="~/Dropbox/gh_projects/pacification_of_the_world/content/problems_with_the_opinion_policy_link.md"}
```

# On globalization
```{r globalization_and_its_discontents, child="~/Dropbox/gh_projects/pacification_of_the_world/content/globalization_and_its_discontents/globalization_and_its_discontents.Rmd"}
```

# The Instrumentalism of Keynesianism
```{r instrumentalism-of-keynesianism, child="~/Dropbox/gh_projects/pacification_of_the_world/content/instrumentalism_of_keynesianism.md"}
```

# The ideology of "citizen engagement"
```{r the_ideology_of_citizen_engagement, child="~/Dropbox/gh_projects/pacification_of_the_world/content/ideology_of_citizen_engagement.md"}
```

# Globalizing co-optation
```{r globalizing_cooptation, child="~/Dropbox/gh_projects/pacification_of_the_world/content/globalizing_cooptation.md"}
```



\pagebreak

# References




